---
title: "R Notebook"
author: "Emmanuel Czar B. Pascua"
output: html_notebook
---


# **IBM HR Analysis**
#### A data exploration case study 
#
#
#
#### **First, let's install some useful packages**

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


```{r message=FALSE, warning=FALSE}
install.packages("tidyverse")
install.packages("janitor")
install.packages("here")
install.packages("skimr")
library(tidyverse)
library(janitor)
library(here)
library(skimr)
```

#### **Let's now load our downloaded csv file from kaggle.com**
```{r}
setwd("D:/FILES/DATA ANALYTICS/PORTFOLIO/IBM HR analytics")
ibm_hr <- read_csv("HR_Employee_Attrition.csv")
head(ibm_hr)
```

#### **Let's now clean our data.**
Check for duplicates. This code will return True if unique employee numbers are equal to total number of rows
```{r}
length(unique(ibm_hr$EmployeeNumber)) == nrow(ibm_hr)
```
Check for missing values
```{r}
valmissing <- na.omit(ibm_hr)

if(nrow(valmissing) == nrow(ibm_hr)){
print("There are 0 missing values in the data")
}
```

It looks like we don't have null values since we still have the original number of rows.

#### **Find correlation**
Try to find correlation in our data set but first we have to drop non-numeric columns including arbitrary and non ordinal data such as EmployeeNumber and EmployeeCount. 

```{r}
ibm_hr_num <- ibm_hr[sapply(ibm_hr, is.numeric)]

ibm_hr_nl <- subset(ibm_hr_num, select = -c(EmployeeNumber,EmployeeCount))
head(ibm_hr_nl)
```
```{r}
install.packages("ggcorrplot")
library(ggcorrplot)
data(ibm_hr_nl)
corr <- round(cor(ibm_hr_nl), 1)
ggcorrplot(corr, legend.title = "Correlation Legend")
```

#### **Let's now use ggplot2 package for the visualization to check for attrition rate across different variables**

```{r}
ggplot(ibm_hr, aes(x = Attrition)) +
  geom_bar()
```


```{r}
#Generate visuals to compare attrition using different variables 
ggplot(ibm_hr, aes(Department, fill=Attrition))+ geom_bar(position="dodge2")
ggplot(ibm_hr, aes(BusinessTravel, fill=Attrition))+ geom_bar(position="dodge2")
ggplot(ibm_hr, aes(EducationField, fill=Attrition))+ geom_bar(position="dodge2")
ggplot(ibm_hr, aes(DistanceFromHome, fill=Attrition))+ geom_bar(position="dodge2")
ggplot(ibm_hr, aes(EnvironmentSatisfaction, fill=Attrition))+ geom_bar(position="dodge2")
ggplot(ibm_hr, aes(Gender, fill=Attrition))+ geom_bar(position="dodge2")
```
#### **Some observations on the visuals**

1. *Department* : Highest attrition in R&D, lowest in HR.
2. *Business Travel*: Highest attrition are those who travel rarely
3. *Education Field*: Highest attrition rate are those in life sciences but could be because LS make up majority of all employees.
4. *Distance from home*: no observed trend
5. *Environment satisfaction*: as expected, highest number of attrition came from those who rated poorly
6. *Gender*: Male employees left the make up the majority of the attrition count but could be because male dominates the majority of the total sample






